---
title: "Hands-on Exercise 1: Geospatial Data Wrangling with R"
editor: visual
---

## Overview

In this hands-on exercise, I learn how to import and wrangling geospatial data using appropriate R packages.

## Getting Started

The code chunk below install and load [sf](https://r-spatial.github.io/sf/) and tidyverse packages into R environment.

```{r}
pacman::p_load(sf, tidyverse, tmap)
```

## Importing Geospatial Data

### Importing polygon feature data in shapefile format

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

### Importing polygon feature data in shapefile form

```{r}
cyclingpath <- st_read(dsn = "data/geospatial", layer = "CyclingPathGazette")
```

### Importing GIS data in kml format

```{r}
preschool <- st_read(dsn = "data/geospatial/PreSchoolsLocation.kml")
```

## Checking the Content of A Simple Feature Data Frame

### Working with st_geometry

```{r}
st_geometry(mpsz)
```

### Working with glimpse

```{r}
glimpse(mpsz)
```

### Working with head()

```{r}
head(mpsz, n=5)
```

## Plotting the Geospatial Data

### Plot Everything

```{r}
plot(mpsz, max.plot = 15)
```

### Plot only the Geometry

```{r}
plot(st_geometry(mpsz))
```

### Plot Based on Specific Object

```{r}
plot(mpsz["PLN_AREA_N"])
```

## Working with Projection

### Assigning EPSG code to a simple feature data frame

```{r}
st_crs(mpsz)
```

assign correct EPSG code to mspz

```{r}
mpsz3414 <- st_set_crs(mpsz, 3414)
```

check the correction result

```{r}
st_crs(mpsz3414)
```

### Transforming the projection of preschool from wgs84 to svy21

```{r}
preschool3414 <- st_transform(preschool, crs = 3414)
```

## Importing and Converting an Aspatial Data

### Importing the aspatial data

```{r}
listings <- read_csv("data/aspatial/listings.csv")
```

show the listing

```{r}
list(listings)
```

### Creating a simple feature data frame from an aspatial dara frame

convert listing into simple feature

```{r}
listings_sf <- st_as_sf(listings, coords = c("longitude", "latitude"), crs=4326) %>% st_transform(crs = 3414)
```

examine the content

```{r}
glimpse(listings_sf)
```

## Geoprocessing with sf package

### Buffering

compute 5-meter buffers around cycling paths

```{r}
buffer_cycling <- st_buffer(cyclingpath, dist=5, nQuadSegs = 30)
```

calculate the area of buffers

```{r}
buffer_cycling$AREA <- st_area(buffer_cycling)
```

derive the total land involved

```{r}
sum(buffer_cycling$AREA)
```

### Point-in-polygon count

identify pre-schools located inside each Planning Subzone, then calculate numbers of pre-schools inside each planning subzone

```{r}
mpsz3414$`PreSch Count`<- lengths(st_intersects(mpsz3414, preschool3414))
```

check the summary statistics

```{r}
summary(mpsz3414$`PreSch Count`)
```

check the subzone with most number of pre-school

```{r}
top_n(mpsz3414, 1, `PreSch Count`)
```

Calculate the density of pre-school by planning subzone. First step, derive the area of each planning subzone.

```{r}
mpsz3414$Area <- mpsz3414 %>%
  st_area()
```

Second step, compute density

```{r}
mpsz3414 <- mpsz3414 %>%
  mutate(`PreSch Density` = `PreSch Count`/Area * 1000000)
```

## Exploratory Data Analysis (EDA)

Plotting hitogram for PreSch Density

```{r}
hist(mpsz3414$`PreSch Density`)
```

Plot better histogram using GG Plot:

```{r}
ggplot(data=mpsz3414, 
       aes(x= as.numeric(`PreSch Density`)))+
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  labs(title = "Are pre-school even distributed in Singapore?",
       subtitle= "There are many planning sub-zones with a single pre-school, on the other hand, \nthere are two planning sub-zones with at least 20 pre-schools",
      x = "Pre-school density (per km sq)",
      y = "Frequency")
```

Plot a scatterplot using GGPlot

```{r}
ggplot(data=mpsz3414, 
       aes(y = `PreSch Count`, 
           x= as.numeric(`PreSch Density`)))+
  geom_point(color="black", 
             fill="light blue") +
  xlim(0, 40) +
  ylim(0, 40) +
  labs(title = "",
      x = "Pre-school density (per km sq)",
      y = "Pre-school count")
```