---
title: "Hands-on Exercise 1: Choropleth Mapping"
editor: visual
---

## Overview

In this hands-on exercise, we will continue on Exploratory Data Analysis, specifically using Choropleth Mapping

## Import The Libraries

The code chunk below install and load [sf](https://r-spatial.github.io/sf/) and tidyverse packages into R environment.

```{r}
pacman::p_load(sf, tidyverse, tmap)
```

## Importing The Data

### Geospatial Data

```{r}
mpsz <- st_read(dsn = "../data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

```{r}
cyclingpath <- st_read(dsn = "../data/geospatial", layer = "CyclingPathGazette")
```

```{r}
preschool <- st_read(dsn = "../data/geospatial/PreSchoolsLocation.kml")
```

### Attribute Data

```{r}
popdata <- read_csv("../data/aspatial/respopagesextod2011to2020.csv.gz")
```

## Data Preparation

### Data Wrangling

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup()%>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
glimpse(popdata2020)
```

### Joining the attribute data and geospatial data

standardize case of the strings

```{r}
popdata2020 <- popdata2020 %>%
  mutate_at(.vars = vars(PA, SZ), 
          .funs = funs(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
glimpse(popdata2020)
```

join the geographical data and attribute table using planning sub zone as the common identifier

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
glimpse(mpsz_pop2020)
```

write R object to R Data Serialization Format

```{r}
write_rds(mpsz_pop2020, "../data/rds/mpszpop2020.rds")
```

## Choropleth Mapping Geospatial Data using tmap

### Plotting choropleth map using qtm

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, fill = "DEPENDENCY")
```

### Creating a choropleth map by using tmap's elements

for a better quality cartographic choropleth map, use tmap's drawing elements like this

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "Dependency ratio") +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

#### Drawing a base map

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

#### Drawing a choropleth map using tm_polygons

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY")
```

#### Drawing Choropleth map using tm_fill and tm_border

without tm_border, there will be no boundary

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```

add boundary of the planning_subzones by using tm_borders

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY") +
  tm_borders(lwd = 0.1,  alpha = 1)
```

Notice that light-gray border lines have been added on the choropleth map.

The *alpha* argument is used to define transparency number between 0 (totally transparent) and 1 (not transparent). By default, the alpha value of the col is used (normally 1).

Beside *alpha* argument, there are three other arguments for *tm_borders()*, they are:

-   *col* = border colour,

-   *lwd* = border line width. The default is 1, and

-   *lty* = border line type. The default is "solid".

### Data Classification methods of tmap

#### Plotting choropleth maps with built-in classification methods

classify with quantile data with 5 classes

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

Using equal range data classification, the map is not too informative as the data is skewed

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5)
```

#### Plotting choropleth map with custome break

show descriptive statistics for break reference

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

plot the map using breaks vector.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(alpha = 0.5)
```

### Colour Scheme

#### Using ColourBrewer Palette

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 6,
          style = "quantile",
          palette = "Blues") +
  tm_borders(alpha = 0.5)
```

Reverse the color shade

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "-Greens") +
  tm_borders(alpha = 0.5)
```

### Map Layouts

### Adding legend

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "jenks", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone \n(Jenks classification)",
            main.title.position = "center",
            main.title.size = 1,
            legend.height = 0.45, 
            legend.width = 0.35,
            legend.outside = FALSE,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

#### Map Style

Testing with classic style

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```

#### Cartographic Furniture

testing change in compass, scale bar, and tm grid

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "No. of persons") +
  tm_layout(main.title = "Distribution of Dependency Ratio \nby planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

reset the tmap setting to default style

```{r}
tmap_style("white")
```

### Drawing Small Multiple Choropleth Maps

#### By assigning multiple values to at least one of the aesthetic arguments

using ncols in tm_fill

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill(c("YOUNG", "AGED"),
          style = "equal", 
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```

state multiple pair of combinations

```{r}
tm_shape(mpsz_pop2020)+ 
  tm_polygons(c("DEPENDENCY","AGED"), style = c("equal", "quantile"), palette = list("Blues","Greens")) + tm_layout(legend.position = c("right", "bottom"))
```

#### By defining a group-by variable in tm_facets

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) + 
  tm_facets(by="REGION_N", 
            free.coords=TRUE, 
            drop.shapes=TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"), 
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

#### By creating multiple stand-alone maps with tmap_arrange

```{r}
youngmap <- tm_shape(mpsz_pop2020) + tm_polygons("YOUNG", style = "quantile", palette = "Blues")
agedmap <- tm_shape(mpsz_pop2020) + tm_polygons("AGED", style = "quantile", palette = "Blues")
tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

### Mapping Spatial Object Meeting a Selection Criterion

choose Central Region as example

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45, 
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```
